\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-19 ResponseToReviews.tex

When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data.
Note that even with this shift, BATCAVE still performs better than MuTect on the validated mutations (although both perform extremely well).
\RNG{The biology of the shift is...}
\end{response}

Related to this, there is probably a minimum number of high-confidence mutations that are required by BATCAVE to produce reliable results. As with the previous point, ideally the batcaver should address or warn the user if the assumptions are violated.
\begin{response}
Our simulation studies suggest that a few hundred high-confidence mutations are enough to reliably infer even a sharply peaked mutation profile (Fig. 3).
We have modified \texttt{batcaver} so that it now emits a warning if there are fewer than 500 high-confidence mutations in a sample.
\end{response}

Lastly, as correctly pointed out by the second reviewer, the sequencing coverage only makes sense in the context of the purity of the samples and this information should be provided to the reader.
\begin{response}
The direct effect of contamination by normal cells on the BATCAVE algorithm is to bias the estimated allele frequencies that are inputs to Eq.~11 for estimating the overall mutation rate.
This bias can be simply corrected given an estimate of tumor purity.
We have thus added a \texttt{--tumor\_purity} option to \texttt{batcaver}.
When this is enabled, the estimated allele frequencies will be appropriately adjusted before applying Eq.~11.
See new text in Methods for details.

We also reanalyzed our real data examples, using purity estimates from the original publications.
\RNG{
The effect of accounting for tumor purity was to...
We have updated the text, Table~1, and Fig.~4E\&F to reflect these new results.
}

\end{response}

@@

We discuss this issue in a new paragraph in Discussion.
\end{response}

\section*{Reviewer 2}

Mannakee et al have developed a tool called BATCAVE for calling mutations from sequencing data. They claim to have used a novel approach of using tumor and site-specific priors for calling mutations. In my opinion authors have done a good job in terms of using a novel approach which harnesses patient/tumor specific information to call mutations. I agree with the authors that this approach was missing in other tools. I accept this study to be published given the authors address few points mentioned below:

1. Does BATCAVE accounts for tumor purity estimates while calling mutations?
\begin{response}
We neglected to account for the effect of tumor purity on mutation rate estimation from Eq.~11.
We have corrected this omission.
Please see our response to the Associate Editor's comment for details.
\end{response}

### 2019-12-19 NAR_DRAFT.tex
Add tumor purity draft responses. Will need to rerun analyses before finalizing.

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
\bkmcomment{We make two simplifying assumptions here.
First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
Second, we assume 100\% tumor purity.
MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
We encourage users to take advantage of the MuTect contamination correction where appropriate.
In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
}
%We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.

@@

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

@@

Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
\rngcomment{Brian: Add purity estimate, if in paper.}
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

>>>>

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
\rngcomment{Note that sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p$.
All estimated allele frequencies are then multiplied by a factor of $1/p$ before being used in Eq.~\ref{eqn:mut_rate}.
}

@@

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
\rngcomment{Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.}
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we \rngcomment{used the XXX samples} and ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
\rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

@@

Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
